---
title: "Part I - EDA and LC"
author: "Sanjeev Gadre"
date: "September 24, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```

Loading the required libraries

```{r libraries}
library(dplyr)
library(ggplot2)
library(ff)
library(ffbase)

```

Loading data

```{r get-data}
train = read.csv.ffdf(file = "../data/train.csv")
test = read.csv.ffdf(file = "../data/test.csv")
n.train = nrow(train)
n.test = nrow(test)
p.train = ncol(train)

```

Loading utility functions

```{r utilitity-functions}
save.ffdf(train, dir = "../RDA/traindir", overwrite = TRUE)
save.ffdf(test, dir = "../RDA/testdir", overwrite = TRUE)
```


EDA - I

```{r eda-1}
colnames.train = colnames(train[,])
print("The column names for the train dataset are:")
colnames.train

print("Sample values from the first column of the train dataset:")
head(train[,1]); tail(train[,1])
print("Sample values from the last column of the train dataset:")
head(train[, 258]); tail(train[, 258])
```

1.  The first column is `id`. The values of this column don't show any discernable patterns.
2.  The last column is `target` and represents the classification lable value {0, 1}
3.  The remaining 256 columns have *goofy* names and need to be replaced with something easier to work with. This change will be made to both the *train* and *test* datasets.
4.  We save the original column names for future reference.
5.  Finally we ascertain if there are any NA values either in the *train* or the *test* subset.


```{r pre-proc-1}
colnames.train.new = colnames.train
for (i in 2:257) {
  colnames.train.new[i] = paste("var_", i-1, sep = "")
}

colnames(train) = colnames.train.new
colnames(test) = colnames.train.new[1:257]

saveRDS(colnames.train, file = "../RDA/colnames.train.orig")

```


